---
title: "An evaluation of ABC and TAC projections for the Bering Sea and Aleutian Island groundfish"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    lightbox: true
    code-fold: true
    code-summary: "Show code"
  pdf:
    toc: true
execute:
  echo: true
  warning: false
  message: false
bibliography: ref.bib
link-citations: true
---

```{r}
library(here)
library(tidyverse)
library(scales)
library(GGally)

theme_set(theme_minimal(base_size = 12))

data_root <- if (file.exists(file.path("..", "data", "BSAI_OFL_ABC_TAC.csv"))) {
  file.path("..", "data")
} else {
  "data"
}

df <- read_csv(file.path(data_root, "BSAI_OFL_ABC_TAC.csv"), show_col_types = FALSE)

#goaraw <- read_csv(file.path(data_root, "GOA_OFL_ABC_TAC.csv"), show_col_types = FALSE)
goaraw <- read_csv(file.path(data_root, "GOA_OFL_ABC_TAC_specs.csv"), show_col_types = FALSE)

mainspp <- c(
  "Pollock",
  "Yellowfin sole",
  "Pacific cod",
  "Atka mackerel",
  "Northern rock sole",
  "Flathead sole",
  "Pacific ocean perch"
)

dfOY <- df %>%
  filter(OY == 1) %>%
  group_by(Year = ProjYear, lag, Species, Order) %>%
  summarise(
    ABC = sum(ABC, na.rm = TRUE),
    OFL = sum(OFL, na.rm = TRUE),
    TAC = sum(TAC, na.rm = TRUE),
    .groups = "drop"
  )

species_order <- dfOY %>% distinct(Species, Order)

main_share <- dfOY %>%
  filter(lag == 1) %>%
  summarise(share = sum(ABC[Species %in% mainspp], na.rm = TRUE) / sum(ABC, na.rm = TRUE)) %>%
  pull(share)

cv_tbl <- dfOY %>%
  filter(lag == 1, Species %in% mainspp) %>%
  group_by(Species) %>%
  summarise(
    cv_ABC = sd(ABC, na.rm = TRUE) / mean(ABC, na.rm = TRUE),
    cv_TAC = sd(TAC, na.rm = TRUE) / mean(TAC, na.rm = TRUE),
    .groups = "drop"
  )

cv_max <- cv_tbl %>% arrange(desc(cv_ABC)) %>% slice(1)
cv_min_tac <- cv_tbl %>% arrange(cv_TAC) %>% slice(1)

goaOY <- goaraw %>%
  filter(OY == 1, Area == "Total") %>%
  group_by(Year = ProjYear, lag, Species) %>%
  summarise(
    ABC = sum(ABC, na.rm = TRUE),
    OFL = sum(OFL, na.rm = TRUE),
    TAC = sum(TAC, na.rm = TRUE),
    .groups = "drop"
  )

goa_main_spp <- goaOY %>%
  filter(lag == 1) %>%
  group_by(Species) %>%
  summarise(mean_ABC = mean(ABC, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_ABC)) %>%
  slice_head(n = 7) %>%
  pull(Species)

goa_cv_tbl <- goaOY %>%
  filter(lag == 1, Species %in% goa_main_spp) %>%
  group_by(Species) %>%
  summarise(
    cv_ABC = if_else(mean(ABC, na.rm = TRUE) > 0, sd(ABC, na.rm = TRUE) / mean(ABC, na.rm = TRUE), NA_real_),
    cv_TAC = if_else(mean(TAC, na.rm = TRUE) > 0, sd(TAC, na.rm = TRUE) / mean(TAC, na.rm = TRUE), NA_real_),
    .groups = "drop"
  )

goa_cv_max <- goa_cv_tbl %>% arrange(desc(cv_ABC)) %>% slice(1)
goa_cv_min_tac <- goa_cv_tbl %>% arrange(cv_TAC) %>% slice(1)

dat <- dfOY %>%
  filter(Species %in% mainspp) %>%
  mutate(
    Stock = fct_reorder(Species, -Order),
    Species = fct_reorder(Species, Order),
    lag = if_else(lag == 2, "two", "one")
  ) %>%
  select(Stock, Species, Year, TAC, ABC, lag, Order) %>%
  pivot_wider(names_from = lag, values_from = c(TAC, ABC)) %>%
  mutate(
    ABC_deltat = (ABC_one - ABC_two) / 1000,
    TAC_deltat = (TAC_one - TAC_two) / 1000,
    ABC_delta = if_else(ABC_two > 0, (ABC_one - ABC_two) / ABC_two, NA_real_),
    TAC_delta = if_else(ABC_two > 0, (TAC_one - TAC_two) / ABC_two, NA_real_),
    TAC_sign = if_else(TAC_delta < 0, "-", "+"),
    ABC_sign = if_else(ABC_delta < 0, "-", "+")
  )

mean_deltas <- dat %>%
  summarise(
    mean_abc = mean(ABC_delta, na.rm = TRUE),
    mean_tac = mean(TAC_delta, na.rm = TRUE)
  )

mean_by_species <- dat %>%
  group_by(Species) %>%
  summarise(
    mean_abc = mean(ABC_delta, na.rm = TRUE),
    mean_tac = mean(TAC_delta, na.rm = TRUE),
    .groups = "drop"
  )

max_abc <- mean_by_species %>% arrange(desc(mean_abc)) %>% slice(1)
min_abc <- mean_by_species %>% arrange(mean_abc) %>% slice(1)
min_tac <- mean_by_species %>% arrange(mean_tac) %>% slice(1)

rollover <- dfOY %>%
  filter(Year > 2000, Species %in% mainspp) %>%
  mutate(application = if_else(lag == 1, "Used", "Projected")) %>%
  select(Year, Species, ABC, TAC, application) %>%
  pivot_wider(names_from = application, values_from = c(ABC, TAC)) %>%
  group_by(Species) %>%
  arrange(Year) %>%
  mutate(
    Projected_ABC = lag(ABC_Used),
    Projected_TAC = lag(TAC_Used)
  ) %>%
  ungroup() %>%
  mutate(
    ABC_delta = if_else(Projected_ABC > 0, (ABC_Used - Projected_ABC) / Projected_ABC, NA_real_),
    TAC_delta = if_else(Projected_TAC > 0, (TAC_Used - Projected_TAC) / Projected_TAC, NA_real_)
  ) %>%
  left_join(species_order, by = "Species")

rollover_means <- rollover %>%
  summarise(
    mean_abc = mean(ABC_delta, na.rm = TRUE),
    mean_tac = mean(TAC_delta, na.rm = TRUE)
  )

model_err <- dat %>%
  filter(Year > 2000) %>%
  mutate(
    ABC_err_model = if_else(ABC_one > 0, abs(ABC_one - ABC_two) / ABC_one, NA_real_),
    TAC_err_model = if_else(TAC_one > 0, abs(TAC_one - TAC_two) / TAC_one, NA_real_)
  ) %>%
  select(Species, Year, ABC_err_model, TAC_err_model)

rollover_err <- rollover %>%
  mutate(
    ABC_err_roll = if_else(ABC_Used > 0, abs(ABC_Used - Projected_ABC) / ABC_Used, NA_real_),
    TAC_err_roll = if_else(TAC_Used > 0, abs(TAC_Used - Projected_TAC) / TAC_Used, NA_real_)
  ) %>%
  select(Species, Year, ABC_err_roll, TAC_err_roll)

comp_err <- model_err %>%
  inner_join(rollover_err, by = c("Species", "Year"))

comp_summary <- comp_err %>%
  summarise(
    model_better_abc = mean(ABC_err_model < ABC_err_roll, na.rm = TRUE),
    model_better_tac = mean(TAC_err_model < TAC_err_roll, na.rm = TRUE),
    mean_abc_model = mean(ABC_err_model, na.rm = TRUE),
    mean_abc_roll = mean(ABC_err_roll, na.rm = TRUE),
    mean_tac_model = mean(TAC_err_model, na.rm = TRUE),
    mean_tac_roll = mean(TAC_err_roll, na.rm = TRUE)
  )

comp_by_species <- comp_err %>%
  group_by(Species) %>%
  summarise(
    mean_abc_model = mean(ABC_err_model, na.rm = TRUE),
    mean_abc_roll = mean(ABC_err_roll, na.rm = TRUE),
    mean_tac_model = mean(TAC_err_model, na.rm = TRUE),
    mean_tac_roll = mean(TAC_err_roll, na.rm = TRUE),
    .groups = "drop"
  )

species_abc_adv <- sum(comp_by_species$mean_abc_model < comp_by_species$mean_abc_roll, na.rm = TRUE)
species_tac_adv <- sum(comp_by_species$mean_tac_model < comp_by_species$mean_tac_roll, na.rm = TRUE)
```

# Introduction

Annual ABC and TAC decisions in the Bering Sea and Aleutian Islands (BSAI) groundfish fisheries set the management ceilings that balance conservation, stability, and economic planning. Multi-year advice provides predictable expectations, but it is built on forecasts that can change as new data arrive, which can create abrupt revisions in advice.

This work evaluates whether two-year model projections provide a meaningful improvement over a simple rollover of the previous year's values. Quantifying when projections track final outcomes, and when they do not, helps interpret the reliability of advice, supports communication with stakeholders, and identifies where interim updates may add value.

This document compares final (lag 1) ABC and TAC values to their two-year projections (lag 2) using OY = 1 records. The seven main species listed above account for about `r percent(main_share, accuracy = 1)` of total ABC, so most of the signal comes from them.

# Methods

Data were compiled for the 1986–2025 period for the Bering Sea and Aleutian Islands (BSAI) region, consistent with the management framework described in the [BSAI fishery management plan](https://files.npfmc.org/fmp/BSAI/BSAIfmp.pdf) and the [BSAI groundfish amendments action summary](https://files.npfmc.org/fmp/BSAI/BSAIGFAmActionSumm.pdf). The dataset includes ABCs, TACs, and OFLs set by management for the next year and the year after (lag 1 and lag 2). We first evaluated overall variability across the main stocks for both ABC and TAC, then examined patterns in the Pollock stock in more detail.

As an exploratory extension, we attempted to replace the ATTACH catch function with a DSEM-based approach that maps ABC inputs to expected TACs for key species while modeling cross-species dependence and temporal dynamics. This provides a direct comparison to the static SUR-style framework in ATTACH and helps assess whether a dynamic model can better replicate management outcomes.

## Performance measures

For each species $s$ and projection year $t$, let $X_{t,1}$ be the final (lag 1) value and $X_{t,2}$ the two-year projection (lag 2), where $X$ is either ABC or TAC. We computed percent differences as:

$$
\Delta X_{t} = \frac{X_{t,1} - X_{t,2}}{X_{t,2}}.
$$

To keep ABC and TAC deltas on a common scale, TAC percent changes were scaled by the two-year ABC:

$$
\Delta TAC_{t} = \frac{TAC_{t,1} - TAC_{t,2}}{ABC_{t,2}}.
$$

Absolute changes in thousand tons are:

$$
\Delta X_{t}^{(kt)} = \frac{X_{t,1} - X_{t,2}}{1000}.
$$

Interannual variability was summarized using the coefficient of variation:

$$
CV(X) = \frac{sd(X)}{mean(X)}.
$$

For the projection-versus-rollover comparison, we used absolute percent error against the final value:

$$
APE_{model}(X) = \frac{\lvert X_{t,1} - X_{t,2} \rvert}{X_{t,1}}, \quad
APE_{roll}(X) = \frac{\lvert X_{t,1} - X_{t-1,1} \rvert}{X_{t,1}}.
$$


## Dynamic structural equation model setup

This section sets up dynamic structural equation models (DSEMs) using the `dsem` package. DSEMs blend structural equation modeling with state-space time-series dynamics to estimate both contemporaneous effects and temporal persistence, allowing mechanistic interpretation of multivariate ecological systems [@Thorson2024DSEM]. Below, the DSEM work is shown separately for BSAI and GOA. In both regions, the dependent variables are TACs for Pollock and Pacific cod, with contemporaneous ABC inputs drawn from a region-specific set of major species.

Seemingly unrelated regression (SUR) provides a useful static reference point for understanding dynamic structural equation models (DSEM). In SUR, multiple regression equations are estimated jointly, allowing for contemporaneous correlation among equation-specific error terms while excluding any dynamic feedback or causal coupling across outcomes [@Zellner1962SUR]. DSEM can be viewed as a strict generalization of this framework: correlated disturbances are retained, but are embedded within a dynamic state-space formulation that allows for autoregressive behavior, latent processes, and explicit structural pathways across variables and time [@Asparouhov2018DSEM]. From this perspective, SUR corresponds to a single-time-slice, purely contemporaneous special case of DSEM in which temporal dependence and structural effects are suppressed. This connection clarifies how joint modeling gains efficiency through shared stochastic components, while DSEM extends that logic by separating persistent process variation from transient noise and propagating information forward through time.

In the BSAI context, the ATTACH model implemented in the `catchfunction` package uses a two-step regression workflow (ABC → TAC, then TAC → catch) with SUR linking TAC-stage errors across species, and an ensemble of alternative error structures in the catch stage to reflect correlated shocks and the ecosystem cap [@FaigHaynie2020ATTACH]. This provides a static, cross-equation baseline that motivates the move to DSEM when temporal dynamics and lagged effects are of interest.

For clarity and understanding, we constructed similar log-log TAC~ABC regressions for the main stocks and pooled "Others". Here, TACs for each stock were estimated statistically using a log-linear model. For $j = 1, 2, \ldots, n$ stocks, the general model for stock $i$ took the form:

$$
\ln(\mathrm{TAC}_{i,t}) = \alpha_i + \beta_i \ln(\mathrm{ABC}_{i,t}) + \sum_{j \ne i}^{n} \beta_{ij} \ln(\mathrm{ABC}_{j,t}) + \sum_{k=1}^{m} \beta_k I_{k,t} + \varepsilon_{i,t}.
$$

where $\alpha_i$ is the stock-specific intercept for species $i$, $\beta_i$ is the elasticity of the TAC of species $i$ with respect to its own ABC, and $\beta_{ij}$ is the elasticity for the ABC of species $j \neq i$ in the TAC equation for species $i$. The effect ($\beta_k$) of $k = 1, 2, \ldots, m$ events or policy changes (e.g., changes in management, area closures, or implementation of catch share programs) on TAC was also estimated, where $I_{k,t}$ is an indicator variable for event $k$ in year $t$, and $\varepsilon_{i,t}$ denotes the residual error for the prediction in year $t$. 


# Results

## Patterns in catch advice and limits

### BSAI totals by year

Here we show simple annual BSAI totals for TAC, ABC, and OFL (lag 1, OY = 1), summed across species.

```{r}
#| label: fig-bsai-totals
#| fig-cap: "BSAI totals by year (sum across species): TAC, ABC, and OFL."
#| fig-height: 4.5
#| fig-width: 7

bsaiyr <- dfOY %>%
  filter(lag == 1) %>%
  group_by(Year) %>%
  summarise(
    TAC = sum(TAC, na.rm = TRUE),
    ABC = sum(ABC, na.rm = TRUE),
    OFL = sum(OFL, na.rm = TRUE),
    n_TAC = sum(!is.na(TAC)),
    n_ABC = sum(!is.na(ABC)),
    n_OFL = sum(!is.na(OFL)),
    .groups = "drop"
  )

plot_bsai <- bsaiyr %>%
  select(Year, TAC, ABC, OFL) %>%
  pivot_longer(-Year, names_to = "Measure", values_to = "value") %>%
  mutate(
    value = na_if(value, 0),
    Measure = factor(Measure, levels = c("TAC", "ABC", "OFL"))
  )

ggplot(plot_bsai, aes(x = Year, y = value / 1000, color = Measure)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "thousand t", color = NULL) +
  theme(legend.position = "top")
```

### BSAI Pollock

Pollock provides a clear example of how ABC, OFL, and TAC move together over time (see @fig-pollock-ts).
OFL generally sits above ABC, and TAC sits below ABC.

```{r}
#| label: fig-pollock-ts
#| fig-cap: "BSAI Pollock ABC, OFL, and TAC over time (lag 1, OY = 1)."
pollock_ts <- dfOY %>%
  filter(lag == 1, Species == "Pollock") %>%
  pivot_longer(cols = c(ABC, OFL, TAC), names_to = "Measure", values_to = "value") %>%
  mutate(value = na_if(value, 0))

ggplot(pollock_ts, aes(x = Year, y = value / 1000, color = Measure)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = NULL, y = "thousand t", color = NULL)
```

### GOA totals by year

The GOA harvest specifications file was munged into the same long format as the BSAI data.
Here we show annual totals from GOA `Area == "Total"` rows for TAC, ABC, and OFL.
Because the GOA spreadsheet contains some `n/a` fields (especially for OFL/ABC in early years),
we report totals as the sum of available numeric entries.

```{r}
#| label: fig-goa-totals
#| fig-cap: "GOA totals by year (sum across all rows): TAC, ABC, and OFL."
#| fig-height: 4.5
#| fig-width: 7

goayr <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total") %>%
  group_by(Year = ProjYear) %>%
  summarise(
    TAC = sum(TAC, na.rm = TRUE),
    ABC = sum(ABC, na.rm = TRUE),
    OFL = sum(OFL, na.rm = TRUE),
    n_TAC = sum(!is.na(TAC)),
    n_ABC = sum(!is.na(ABC)),
    n_OFL = sum(!is.na(OFL)),
    .groups = "drop"
  )

plot_goa <- goayr %>%
  select(Year, TAC, ABC, OFL) %>%
  pivot_longer(-Year, names_to = "Measure", values_to = "value") %>%
  mutate(
    value = na_if(value, 0),
    Measure = factor(Measure, levels = c("TAC", "ABC", "OFL"))
  )

ggplot(plot_goa, aes(x = Year, y = value / 1000, color = Measure)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "thousand t", color = NULL) +
  theme(legend.position = "top")
```

### GOA Pollock

GOA Pollock follows the same comparison format as Figure 4 (@fig-goa-pollock-ts), showing annual ABC, OFL, and TAC trajectories.

```{r}
#| label: fig-goa-pollock-ts
#| fig-cap: "GOA Pollock ABC, OFL, and TAC over time (lag 1, OY = 1, Area = Total)."
goa_pollock_ts <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total", Species == "Pollock") %>%
  group_by(Year = ProjYear) %>%
  summarise(
    ABC = sum(ABC, na.rm = TRUE),
    OFL = sum(OFL, na.rm = TRUE),
    TAC = sum(TAC, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(ABC, OFL, TAC), names_to = "Measure", values_to = "value") %>%
  mutate(value = na_if(value, 0))

ggplot(goa_pollock_ts, aes(x = Year, y = value / 1000, color = Measure)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = NULL, y = "thousand t", color = NULL)
```

## Variability across species

### BSAI

@fig-cv shows the interannual CVs for ABC and TAC.
Pacific ocean perch has the highest ABC variability (CV about `r round(cv_max$cv_ABC, 2)`), while Pollock has the lowest TAC variability (CV about `r round(cv_min_tac$cv_TAC, 2)`), indicating tighter management consistency for the largest stock.

```{r}
#| label: fig-cv
#| fig-cap: "Interannual variability (CV) of ABC and TAC for the main BSAI species, 1986-2026."
cv_long <- cv_tbl %>%
  pivot_longer(cols = c(cv_ABC, cv_TAC), names_to = "Measure", values_to = "CV") %>%
  mutate(
    Measure = recode(Measure, cv_ABC = "ABC", cv_TAC = "TAC"),
    Species = fct_reorder(Species, CV)
  )

ggplot(cv_long, aes(x = Species, y = CV, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  coord_flip() +
  labs(x = NULL, y = "Coefficient of variation", fill = NULL) +
  theme(legend.position = "top")
```

### GOA

For GOA, we calculated the same CV metrics for the seven largest species by mean ABC (lag 1, OY = 1, `Area == "Total"`).
@fig-cv-goa shows that `r goa_cv_max$Species` has the highest ABC variability (CV about `r round(goa_cv_max$cv_ABC, 2)`), while `r goa_cv_min_tac$Species` has the lowest TAC variability (CV about `r round(goa_cv_min_tac$cv_TAC, 2)`).

```{r}
#| label: fig-cv-goa
#| fig-cap: "Interannual variability (CV) of ABC and TAC for selected GOA species, 1986-2024 (top seven by mean ABC; lag 1, OY = 1, Area = Total)."
goa_cv_long <- goa_cv_tbl %>%
  pivot_longer(cols = c(cv_ABC, cv_TAC), names_to = "Measure", values_to = "CV") %>%
  mutate(
    Measure = recode(Measure, cv_ABC = "ABC", cv_TAC = "TAC"),
    Species = fct_reorder(Species, CV)
  )

ggplot(goa_cv_long, aes(x = Species, y = CV, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  coord_flip() +
  labs(x = NULL, y = "Coefficient of variation", fill = NULL) +
  theme(legend.position = "top")
```

## Two-year vs final differences (percent)

@fig-delta-pct summarizes the percent difference between final and two-year values, scaled by the two-year ABC to keep ABC and TAC deltas on the same scale.
Across species, the average ABC adjustment is about `r percent(mean_deltas$mean_abc, accuracy = 0.1)`, while TAC adjustments are smaller at `r percent(mean_deltas$mean_tac, accuracy = 0.1)`.
Atka mackerel has the largest positive mean ABC revision (`r percent(max_abc$mean_abc, accuracy = 0.1)`), while Northern rock sole shows the only negative mean ABC adjustment (`r percent(min_abc$mean_abc, accuracy = 0.1)`); Pollock has the most negative mean TAC change (`r percent(min_tac$mean_tac, accuracy = 0.1)`).

```{r}
#| label: fig-delta-pct
#| fig-cap: "Percent changes between final (lag 1) and two-year (lag 2) values, scaled by the two-year ABC."
#| fig-height: 7
#| fig-width: 8
delta_long_pct <- dat %>%
  filter(Year > 1990) %>%
  select(Year, Species, Order, ABC_delta, TAC_delta) %>%
  pivot_longer(cols = c(ABC_delta, TAC_delta), names_to = "Measure", values_to = "Delta") %>%
  mutate(
    Measure = recode(Measure, ABC_delta = "ABC", TAC_delta = "TAC"),
    Species = fct_reorder(Species, Order)
  )

ggplot(delta_long_pct, aes(x = Year, y = Delta, fill = Delta)) +
  geom_col() +
  facet_grid(Species ~ Measure, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_gradient2(
    low = muted("red"),
    mid = "white",
    high = muted("blue"),
    midpoint = 0,
    na.value = "grey90"
  ) +
  labs(x = NULL, y = "Percent change", fill = "Change") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )
```

## Two-year vs final differences (kt)

The heatmaps in @fig-delta-kt-heat highlight the absolute magnitude of adjustments in thousands of tons.
Large swings cluster in the biggest stocks (notably Pollock), while smaller species show less absolute movement even when percent changes are sizable.

```{r}
#| label: fig-delta-kt-heat
#| fig-cap: "Absolute changes between final and two-year values (thousand tons)."
#| fig-height: 7
#| fig-width: 8
complete_years <- dat %>%
  group_by(Year) %>%
  summarise(miss = sum(is.na(ABC_deltat) | is.na(TAC_deltat)), .groups = "drop") %>%
  filter(miss == 0) %>%
  pull(Year)

delta_long_kt <- dat %>%
  filter(Year > 1990, Year %in% complete_years) %>%
  select(Year, Species, Order, ABC_deltat, TAC_deltat) %>%
  pivot_longer(cols = c(ABC_deltat, TAC_deltat), names_to = "Measure", values_to = "Delta") %>%
  mutate(
    Measure = recode(Measure, ABC_deltat = "ABC", TAC_deltat = "TAC"),
    Species = fct_reorder(Species, Order)
  )

ggplot(delta_long_kt, aes(x = Year, y = Species, fill = Delta)) +
  geom_tile() +
  facet_grid(Measure ~ .) +
  scale_fill_gradient2(
    low = muted("red"),
    mid = "white",
    high = muted("blue"),
    midpoint = 0,
    na.value = "grey90",
    labels = comma
  ) +
  labs(x = NULL, y = NULL, fill = "Change (kt)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The results shown in
@fig-delta-kt-heat-nopollock is provided to show contrast among other stocks with pollock removed.

```{r}
#| label: fig-delta-kt-heat-nopollock
#| fig-cap: "Absolute changes between final and two-year values (thousand tons), excluding Pollock."
#| fig-height: 7
#| fig-width: 8
delta_long_kt_nopollock <- delta_long_kt %>%
  filter(Species != "Pollock")

ggplot(delta_long_kt_nopollock, aes(x = Year, y = Species, fill = Delta)) +
  geom_tile() +
  facet_grid(Measure ~ .) +
  scale_fill_gradient2(
    low = muted("red"),
    mid = "white",
    high = muted("blue"),
    midpoint = 0,
    na.value = "grey90",
    labels = comma
  ) +
  labs(x = NULL, y = NULL, fill = "Change (kt)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Rollover scenario

To gauge a simple alternative, @fig-rollover compares final values to a naive rollover that uses the previous year's final value as the "projection."
The average rollover changes are small but positive (`r percent(rollover_means$mean_abc, accuracy = 0.1)` for ABC and `r percent(rollover_means$mean_tac, accuracy = 0.1)` for TAC), suggesting a mild upward drift when viewed against a strict carry-forward baseline.

```{r}
#| label: fig-rollover
#| fig-cap: "Rollover comparison: final values vs prior-year final values used as a naive projection."
#| fig-height: 7
#| fig-width: 8
rollover_long <- rollover %>%
  select(Year, Species, Order, ABC_delta, TAC_delta) %>%
  pivot_longer(cols = c(ABC_delta, TAC_delta), names_to = "Measure", values_to = "Delta") %>%
  mutate(
    Measure = recode(Measure, ABC_delta = "ABC", TAC_delta = "TAC"),
    Species = fct_reorder(Species, Order)
  )

ggplot(rollover_long, aes(x = Year, y = Delta, fill = Delta)) +
  geom_col() +
  facet_grid(Species ~ Measure, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_gradient2(
    low = muted("red"),
    mid = "white",
    high = muted("blue"),
    midpoint = 0,
    na.value = "grey90"
  ) +
  labs(x = NULL, y = "Rollover change", fill = "Change") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  ) + ggthemes::theme_few()
```

## Projection vs rollover accuracy

Using the overlap period (2001 onward), we compared how close the two-year model projections (lag 2) and a naive rollover (previous year's lag 1 value) came to the final values used (lag 1).
On average, the model-based projections show slightly lower absolute percent error for ABC (`r percent(comp_summary$mean_abc_model, accuracy = 0.1)` vs `r percent(comp_summary$mean_abc_roll, accuracy = 0.1)`), and a similarly small improvement for TAC (`r percent(comp_summary$mean_tac_model, accuracy = 0.1)` vs `r percent(comp_summary$mean_tac_roll, accuracy = 0.1)`).
However, the model is closer than the rollover in only `r percent(comp_summary$model_better_abc, accuracy = 1)` of ABC cases and `r percent(comp_summary$model_better_tac, accuracy = 1)` of TAC cases, and the mean advantage appears in `r species_abc_adv` of 7 stocks for ABC and `r species_tac_adv` of 7 stocks for TAC.
This suggests there is enough information to compare the two approaches, but the performance edge of two-year projections over a rollover baseline is modest and uneven—more evident for ABC than for TAC.
@tbl-proj-roll summarizes the mean absolute percent errors by species for each approach.

```{r}
#| label: tbl-proj-roll
#| tbl-cap: "Mean absolute percent error for two-year projections vs rollover baselines (2001+)."
tbl_proj_roll <- comp_by_species %>%
  mutate(
    abc_diff = mean_abc_model - mean_abc_roll,
    tac_diff = mean_tac_model - mean_tac_roll
  ) %>%
  select(
    Species,
    mean_abc_model,
    mean_abc_roll,
    abc_diff,
    mean_tac_model,
    mean_tac_roll,
    tac_diff
  ) %>%
  mutate(across(-Species, ~ percent(.x, accuracy = 0.1)))

knitr::kable(
  tbl_proj_roll,
  col.names = c(
    "Species",
    "ABC model",
    "ABC rollover",
    "ABC model - rollover",
    "TAC model",
    "TAC rollover",
    "TAC model - rollover"
  ),
  align = "lrrrrrr"
)
```

## Projected vs used time series

@fig-ts contrasts the two-year projections with the final values used in the next year.
The projected series are generally smoother, with the largest departures in higher-variability 
stocks (Pacific ocean perch and Flathead sole), which aligns with the elevated CVs in @fig-cv.

```{r}
#| label: fig-ts
#| fig-cap: "Projected (lag 2) vs used (lag 1) ABC time series by species."
ts_dat <- dfOY %>%
  filter(Year > 2004, Species %in% mainspp) %>%
  mutate(Series = if_else(lag == 1, "Used (lag 1)", "Projected (lag 2)")) %>%
  select(Year, Species, ABC, Series) %>%
  left_join(species_order, by = "Species") %>%
  mutate(Species = fct_reorder(Species, Order))

ggplot(ts_dat, aes(x = Year, y = ABC / 1000, color = Series)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~Species, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = NULL, y = "ABC (thousand t)", color = NULL)
```

## TAC cross-species relationships

### BSAI

@fig-tac-pairs shows pairwise relationships among BSAI TAC values for the main species, with all 
remaining species pooled as "Others," using points with a smooth fit in the lower triangle.

```{r}
#| label: fig-tac-pairs
#| fig-cap: "BSAI pairs plot of TAC across the main species plus Others (lag 1, OY = 1)."
tac_pairs_spp <- setdiff(mainspp, c("Pacific ocean perch", "Flathead sole"))

tac_main <- dfOY %>%
  filter(lag == 1, Species %in% tac_pairs_spp) %>%
  mutate(TAC = TAC / 1000) %>%
  select(Year, Species, TAC) %>%
  pivot_wider(names_from = Species, values_from = TAC)

tac_other <- dfOY %>%
  filter(lag == 1, !Species %in% tac_pairs_spp) %>%
  group_by(Year) %>%
  summarise(Others = sum(TAC, na.rm = TRUE) / 1000, .groups = "drop")

tac_wide <- tac_main %>%
  left_join(tac_other, by = "Year") %>%
  select(Year, all_of(tac_pairs_spp), Others)

spp_labels <- c(tac_pairs_spp, "Others")
spp_cols <- gsub(" ", "_", spp_labels)
tac_wide <- tac_wide %>% rename_with(~ spp_cols, all_of(spp_labels))

# Draw lower-triangle pairs panels with points plus a LOESS trend.
lower_points_smooth <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_point(alpha = 0.5, size = 0.7) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.6, color = "steelblue")
}

GGally::ggpairs(
  tac_wide,
  columns = spp_cols,
  columnLabels = spp_labels,
  lower = list(continuous = lower_points_smooth),
  diag = list(continuous = "densityDiag"),
  upper = list(continuous = "cor")
) +
  ggthemes::theme_few(base_size = 9)
```

### GOA

@fig-goa-tac-pairs shows the analogous GOA TAC relationships using the top five species by total TAC
(lag 1, OY = 1, `Area == "Total"`), with all remaining GOA species pooled as "Others."

```{r}
#| label: fig-goa-tac-pairs
#| fig-cap: "GOA pairs plot of TAC across top five species (by total TAC) plus Others (lag 1, OY = 1, Area = Total)."
goa_tac_pairs_spp <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total") %>%
  group_by(Species) %>%
  summarise(total_TAC = sum(TAC, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(total_TAC), total_TAC > 0) %>%
  arrange(desc(total_TAC)) %>%
  slice_head(n = 5) %>%
  pull(Species)

goa_tac_main <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total", Species %in% goa_tac_pairs_spp) %>%
  group_by(Year = ProjYear, Species) %>%
  summarise(TAC = sum(TAC, na.rm = TRUE) / 1000, .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = TAC)

goa_tac_other <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total", !Species %in% goa_tac_pairs_spp) %>%
  group_by(Year = ProjYear) %>%
  summarise(Others = sum(TAC, na.rm = TRUE) / 1000, .groups = "drop")

goa_tac_wide <- goa_tac_main %>%
  left_join(goa_tac_other, by = "Year") %>%
  select(Year, all_of(goa_tac_pairs_spp), Others)

goa_spp_labels <- c(goa_tac_pairs_spp, "Others")
goa_spp_cols <- gsub("[^A-Za-z0-9]+", "_", goa_spp_labels)
goa_tac_wide <- goa_tac_wide %>% rename_with(~ goa_spp_cols, all_of(goa_spp_labels))

GGally::ggpairs(
  goa_tac_wide,
  columns = goa_spp_cols,
  columnLabels = goa_spp_labels,
  lower = list(continuous = lower_points_smooth),
  diag = list(continuous = "densityDiag"),
  upper = list(continuous = "cor")
) +
  ggthemes::theme_few(base_size = 9)
```


## TAC-ABC regressions
As noted, we fit multiple regressions with log(ABC) for the focal stock, separate log(ABC) terms for each other stock group, and a centered year term to proxy policy indicators.

```{r}
# Convert arbitrary stock labels into safe column-name fragments.
safe_names <- function(x) {
  x %>%
    gsub("[^A-Za-z0-9]+", "_", ., perl = TRUE) %>%
    gsub("^_|_$", "", ., perl = TRUE)
}

# Fit stock-specific log(TAC) regressions using own-stock ABC, other-stock ABC,
# and a centered year effect; return coefficient summaries and fitted values.
fit_tac_abc_regressions <- function(data, main_groups) {
  group_names <- c(main_groups, "Other")

  reg_base <- data %>%
    mutate(Group = if_else(Species %in% main_groups, Species, "Other")) %>%
    group_by(Year, Group) %>%
    summarise(
      ABC = sum(ABC, na.rm = TRUE),
      TAC = sum(TAC, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_wider(names_from = Group, values_from = c(ABC, TAC), values_fill = 0)

  names(reg_base) <- safe_names(names(reg_base))
  group_safe <- safe_names(group_names)

  reg_fits <- purrr::map2(group_names, group_safe, function(group, group_safe_i) {
    abc_col <- paste0("ABC_", group_safe_i)
    tac_col <- paste0("TAC_", group_safe_i)
    other_groups <- setdiff(group_names, group)
    other_cols <- paste0("ABC_", safe_names(other_groups))

    df <- reg_base %>%
      transmute(
        Year,
        ABC_i = .data[[abc_col]],
        TAC_i = .data[[tac_col]],
        across(all_of(other_cols), ~ .x, .names = "{.col}"),
        Year_c = Year - mean(Year)
      ) %>%
      filter(TAC_i > 0) %>%
      filter(if_all(all_of(c("ABC_i", other_cols)), ~ .x > 0)) %>%
      mutate(
        log_TAC_i = log(TAC_i),
        log_ABC_i = log(ABC_i),
        across(all_of(other_cols), ~ log(.x), .names = "log_{.col}")
      )

    rhs <- c("log_ABC_i", paste0("log_", other_cols), "Year_c")
    model <- lm(as.formula(paste("log_TAC_i ~", paste(rhs, collapse = " + "))), data = df)
    list(species = group, data = df, model = model)
  })

  reg_results <- purrr::map_dfr(reg_fits, function(x) {
    coefs <- coef(x$model)
    out <- tibble(
      Species = x$species,
      n = nrow(x$data),
      Intercept = unname(coefs["(Intercept)"]),
      log_ABC = unname(coefs["log_ABC_i"]),
      Year_c = unname(coefs["Year_c"]),
      R2 = summary(x$model)$r.squared
    )

    for (g in group_names) {
      col_name <- paste0("log_ABC_", safe_names(g))
      out[[col_name]] <- if (g == x$species) NA_real_ else unname(coefs[col_name])
    }

    out
  }) %>%
    mutate(order = match(Species, group_names)) %>%
    arrange(order) %>%
    select(-order)

  reg_pred <- purrr::map_dfr(reg_fits, function(x) {
    pred_log <- predict(x$model, newdata = x$data)
    tibble(
      Year = x$data$Year,
      Species = x$species,
      Observed = x$data$TAC_i,
      Predicted = exp(pred_log)
    )
  })

  list(
    results = reg_results,
    pred = reg_pred,
    group_names = group_names
  )
}

# Format the regression coefficient summary as a gt table.
render_regression_table <- function(reg_results, group_names) {
  abc_cols <- paste0("log_ABC_", safe_names(group_names))

  reg_results %>%
    gt::gt(rowname_col = "Species") %>%
  gt::cols_label(
    n = "n",
    Intercept = "Intercept",
    log_ABC = "log(ABC)",
    Year_c = "Year (centered)",
    R2 = "R2",
    !!!setNames(paste0("log(ABC ", group_names, ")"), abc_cols)
  ) %>%
  gt::fmt_number(columns = c(Intercept, log_ABC, all_of(abc_cols), Year_c, R2), decimals = 3) %>%
  gt::fmt_number(columns = n, decimals = 0)
}

# Plot observed TAC against fitted regression predictions by stock.
plot_regression_predictions <- function(reg_pred) {
  ggplot(reg_pred, aes(x = Year)) +
    geom_line(aes(y = Predicted, color = "Predicted"), linewidth = 1) +
    geom_point(aes(y = Observed, color = "Observed"), size = 1.2, alpha = 0.8) +
    facet_grid(Species ~ ., scales = "free_y") +
    scale_color_manual(values = c(Predicted = "steelblue", Observed = "grey40")) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "TAC", color = NULL)
}

bsai_reg_main5 <- setdiff(mainspp, c("Pacific ocean perch", "Flathead sole"))
bsai_reg <- fit_tac_abc_regressions(
  data = dfOY %>% filter(lag == 1),
  main_groups = bsai_reg_main5
)

goa_reg_main5 <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total") %>%
  group_by(Species) %>%
  summarise(total_TAC = sum(TAC, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(total_TAC), total_TAC > 0) %>%
  arrange(desc(total_TAC)) %>%
  slice_head(n = 5) %>%
  pull(Species)

goa_reg <- fit_tac_abc_regressions(
  data = goaOY %>% filter(lag == 1),
  main_groups = goa_reg_main5
)
```

### BSAI

For BSAI, we retain the existing five-stock specification: Pollock, Pacific cod, Yellowfin sole, Atka mackerel, Northern rock sole, plus a pooled "Other" category.

```{r}
#| label: tbl-pre-dsem-reg-bsai
#| tbl-cap: "BSAI log-log TAC~ABC regressions with separate other-stock terms (lag 1, OY = 1)."
render_regression_table(bsai_reg$results, bsai_reg$group_names)
```

```{r}
#| label: fig-pre-dsem-pred-bsai
#| fig-cap: "BSAI observed vs predicted TAC from log-linear multiple regressions (lag 1, OY = 1)."
#| fig-height: 5
#| fig-width: 7
plot_regression_predictions(bsai_reg$pred)
```

### GOA

For GOA, we use the top five species by total TAC from the `Area == "Total"` series (`r paste(goa_reg_main5, collapse = ", ")`), plus a pooled "Other" category.

```{r}
#| label: tbl-pre-dsem-reg-goa
#| tbl-cap: "GOA log-log TAC~ABC regressions with separate other-stock terms (lag 1, OY = 1, Area = Total)."
render_regression_table(goa_reg$results, goa_reg$group_names)
```

```{r}
#| label: fig-pre-dsem-pred-goa
#| fig-cap: "GOA observed vs predicted TAC from log-linear multiple regressions (lag 1, OY = 1, Area = Total)."
#| fig-height: 5
#| fig-width: 7
plot_regression_predictions(goa_reg$pred)
```

## GOA OY-cap rescaling alternative

As a GOA-specific alternative, we considered the weighted OY-cap rescaling method described in [the separate GOA OY-cap note](rovellini-goa-oy-cap.qmd). The method starts with projected stock-specific allocations $C_i$ and, when the total exceeds a GOA-wide cap $OY$, rescales them to

$$
C'_i = C_i \times r^{(1 / (w_i m))}, \qquad r = \frac{OY}{\sum_i C_i},
$$

where $w_i > 0$ are stock-specific weights and $m$ is a common multiplier chosen so that the adjusted allocations sum to the cap. Higher weights imply less reduction, so the method can preserve relatively prioritized stocks while still satisfying the aggregate constraint.

For illustration, we applied the method to the latest GOA `Area == "Total"` ABC vector and evaluated three lower cap levels: 95%, 90%, and 85% of the observed aggregate ABC. Because this repository does not include ex-vessel value series or explicit policy weights, we use each stock's historical mean TAC/ABC ratio as a simple proxy weight. That makes this a sensitivity analysis rather than a recommended operational rule, but it is still useful for seeing how an explicit cap-rescaling mechanism differs from the regression and DSEM approaches above.

```{r}
#| label: goa-oy-cap-setup
# Rescale projected catches to satisfy an aggregate OY cap while preserving
# relative priority through stock-specific weights.
rescale_oy_cap <- function(catch, weights, oy_cap, tol = 1e-8, max_iter = 200) {
  catch <- as.numeric(catch)
  weights <- as.numeric(weights)

  total_catch <- sum(catch, na.rm = TRUE)
  if (!is.finite(total_catch) || total_catch <= 0) {
    stop("Total projected catch must be positive.")
  }

  if (oy_cap >= total_catch) {
    return(list(multiplier = NA_real_, scaled = catch, binding = FALSE))
  }

  r <- oy_cap / total_catch
  # Measure the current difference between the rescaled total and the cap.
  total_gap <- function(m) {
    sum(catch * r^(1 / (weights * m))) - oy_cap
  }

  lower <- 1e-6
  upper <- 2
  while (total_gap(upper) < 0 && upper < 1e6) {
    upper <- upper * 2
  }

  for (i in seq_len(max_iter)) {
    mid <- (lower + upper) / 2
    gap_mid <- total_gap(mid)
    if (abs(gap_mid) < tol) break
    if (gap_mid > 0) {
      upper <- mid
    } else {
      lower <- mid
    }
  }

  multiplier <- (lower + upper) / 2
  scaled <- catch * r^(1 / (weights * multiplier))

  list(multiplier = multiplier, scaled = scaled, binding = TRUE)
}

goa_oy_latest_year <- goaOY %>%
  filter(lag == 1) %>%
  summarise(Year = max(Year, na.rm = TRUE)) %>%
  pull(Year)

goa_oy_weights <- goaOY %>%
  filter(lag == 1, is.finite(ABC), ABC > 0, is.finite(TAC), TAC >= 0) %>%
  mutate(weight = TAC / ABC) %>%
  filter(is.finite(weight), weight > 0) %>%
  group_by(Species) %>%
  summarise(weight = mean(weight, na.rm = TRUE), .groups = "drop")

goa_oy_base <- goaOY %>%
  filter(lag == 1, Year == goa_oy_latest_year, is.finite(ABC), ABC > 0) %>%
  select(Species, ABC, TAC) %>%
  left_join(goa_oy_weights, by = "Species") %>%
  mutate(weight = if_else(is.finite(weight) & weight > 0, weight, 0.01)) %>%
  arrange(desc(ABC))

goa_oy_total_abc <- sum(goa_oy_base$ABC, na.rm = TRUE)
goa_oy_total_tac <- sum(goa_oy_base$TAC, na.rm = TRUE)

goa_oy_cap_levels <- tibble(
  Scenario = c("Observed ABC (100%)", "95% cap", "90% cap", "85% cap"),
  cap_factor = c(1.00, 0.95, 0.90, 0.85),
  OY_cap = goa_oy_total_abc * cap_factor
)

goa_oy_alloc <- purrr::map_dfr(seq_len(nrow(goa_oy_cap_levels)), function(i) {
  scenario_row <- goa_oy_cap_levels[i, ]
  fit <- rescale_oy_cap(
    catch = goa_oy_base$ABC,
    weights = goa_oy_base$weight,
    oy_cap = scenario_row$OY_cap
  )

  goa_oy_base %>%
    mutate(
      Scenario = scenario_row$Scenario,
      cap_factor = scenario_row$cap_factor,
      OY_cap = scenario_row$OY_cap,
      multiplier = fit$multiplier,
      ABC_rescaled = fit$scaled,
      pct_reduction = 1 - ABC_rescaled / ABC
    )
})

goa_oy_summary <- goa_oy_alloc %>%
  group_by(Scenario, cap_factor, OY_cap) %>%
  summarise(
    multiplier = first(multiplier),
    total_ABC = sum(ABC, na.rm = TRUE),
    total_rescaled = sum(ABC_rescaled, na.rm = TRUE),
    mean_reduction = mean(pct_reduction, na.rm = TRUE),
    pollock_reduction = pct_reduction[Species == "Pollock"][1],
    pacific_cod_reduction = pct_reduction[Species == "Pacific cod"][1],
    arrowtooth_reduction = pct_reduction[Species == "Arrowtooth Flounder"][1],
    .groups = "drop"
  )

goa_oy_top10 <- goa_oy_base %>%
  slice_head(n = 10) %>%
  pull(Species)
```

The baseline GOA ABC total in `r goa_oy_latest_year` is `r comma(goa_oy_total_abc)` t, compared with aggregate TAC of `r comma(goa_oy_total_tac)` t. Under the weighting rule used here, the cap-rescaling method reduces lower-priority stocks more than high-utilization stocks such as Pollock, Pacific ocean perch, and Rex sole, while still preserving strictly positive allocations for all stocks with positive baseline ABC.

```{r}
#| label: tbl-goa-oy-cap
#| tbl-cap: "Illustrative GOA OY-cap rescaling scenarios for the latest annual ABC vector."
goa_oy_summary %>%
  mutate(
    OY_cap = comma(round(OY_cap)),
    multiplier = if_else(is.na(multiplier), "NA", format(round(multiplier, 3), nsmall = 3)),
    total_ABC = comma(round(total_ABC)),
    total_rescaled = comma(round(total_rescaled)),
    mean_reduction = percent(mean_reduction, accuracy = 0.1),
    pollock_reduction = percent(pollock_reduction, accuracy = 0.1),
    pacific_cod_reduction = percent(pacific_cod_reduction, accuracy = 0.1),
    arrowtooth_reduction = percent(arrowtooth_reduction, accuracy = 0.1)
  ) %>%
  select(
    Scenario,
    OY_cap,
    multiplier,
    total_ABC,
    total_rescaled,
    mean_reduction,
    pollock_reduction,
    pacific_cod_reduction,
    arrowtooth_reduction
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    OY_cap = "OY cap",
    multiplier = "m",
    total_ABC = "Baseline ABC",
    total_rescaled = "Rescaled ABC",
    mean_reduction = "Mean reduction",
    pollock_reduction = "Pollock reduction",
    pacific_cod_reduction = "Pacific cod reduction",
    arrowtooth_reduction = "Arrowtooth reduction"
  )
```

```{r}
#| label: fig-goa-oy-cap
#| fig-cap: "Illustrative GOA OY-cap rescaling for the ten largest stocks in the latest year, under alternative aggregate cap levels."
#| fig-height: 5.5
#| fig-width: 8
goa_oy_plot <- goa_oy_alloc %>%
  filter(cap_factor < 1, Species %in% goa_oy_top10) %>%
  mutate(
    Scenario = factor(Scenario, levels = c("95% cap", "90% cap", "85% cap")),
    Species = factor(Species, levels = rev(goa_oy_top10))
  )

ggplot(goa_oy_plot, aes(x = Species, y = pct_reduction, fill = Scenario)) +
  geom_col(position = position_dodge(width = 0.8)) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Reduction from baseline ABC", fill = NULL) +
  theme(legend.position = "top")
```
## DAGs for the DSEM

```{r}
#| label: dsem-helpers
#| eval: true
library(dsem)

# Convert arbitrary stock labels into safe column-name fragments.
safe_names <- function(x) {
  x %>%
    gsub("[^A-Za-z0-9]+", "_", ., perl = TRUE) %>%
    gsub("^_|_$", "", ., perl = TRUE)
}

# Return vertically centered node positions for DAG plots.
centered_positions <- function(n) {
  if (n <= 1) {
    0
  } else {
    seq(from = (n - 1) / 2, to = -(n - 1) / 2, length.out = n)
  }
}

# Build a static or dynamic DAG showing ABC-to-TAC links for the selected stocks.
plot_dsem_dag <- function(dep_species, indep_species, dynamic = FALSE) {
  dep_nodes <- paste0("TAC_", safe_names(dep_species))
  indep_nodes <- paste0("ABC_", safe_names(indep_species))

  static_formulas <- lapply(dep_nodes, function(dep_node) {
    reformulate(indep_nodes, response = dep_node)
  })

  x_coords <- setNames(rep(0, length(indep_nodes)), indep_nodes)
  y_coords <- setNames(centered_positions(length(indep_nodes)), indep_nodes)

  if (dynamic) {
    lag_nodes <- paste0(dep_nodes, "_lag")
    dynamic_formulas <- Map(function(dep_node, lag_node) {
      reformulate(c(indep_nodes, lag_node), response = dep_node)
    }, dep_nodes, lag_nodes)

    labels <- c(
      setNames(paste0("ABC ", indep_species, " (t)"), indep_nodes),
      setNames(paste0("TAC ", dep_species, " (t-1)"), lag_nodes),
      setNames(paste0("TAC ", dep_species, " (t)"), dep_nodes)
    )

    x_coords <- c(
      x_coords,
      setNames(rep(1, length(lag_nodes)), lag_nodes),
      setNames(rep(2, length(dep_nodes)), dep_nodes)
    )
    dep_y <- centered_positions(length(dep_nodes))
    y_coords <- c(
      y_coords,
      setNames(dep_y, lag_nodes),
      setNames(dep_y, dep_nodes)
    )

    dag_obj <- do.call(
      ggdag::dagify,
      c(dynamic_formulas, list(labels = labels, coords = list(x = x_coords, y = y_coords)))
    )
  } else {
    labels <- c(
      setNames(paste("ABC:", indep_species), indep_nodes),
      setNames(paste("TAC:", dep_species), dep_nodes)
    )

    x_coords <- c(x_coords, setNames(rep(2, length(dep_nodes)), dep_nodes))
    y_coords <- c(y_coords, setNames(centered_positions(length(dep_nodes)), dep_nodes))

    dag_obj <- do.call(
      ggdag::dagify,
      c(static_formulas, list(labels = labels, coords = list(x = x_coords, y = y_coords)))
    )
  }

  dag_tidy <- ggdag::tidy_dagitty(dag_obj)

  ggplot(dag_tidy, aes(x = x, y = y, xend = xend, yend = yend)) +
    ggdag::geom_dag_edges() +
    ggdag::geom_dag_node() +
    ggdag::geom_dag_label_repel(aes(label = label), color = "black", size = 3) +
    ggdag::theme_dag()
}

# dsem has changed allowed values for gmrf_parameterization across versions.
# Choose a value that works for the installed package (GitHub Actions may differ
# from local).
# Find a `gmrf_parameterization` value accepted by the installed dsem version.
choose_gmrf_param <- function() {
  candidates <- c(
    "full", "project", "gmrf_project", "mvn_project", # seen in newer versions
    "separable", "projection"                           # seen in older versions
  )
  for (x in candidates) {
    ok <- tryCatch({
      dsem::dsem_control(gmrf_parameterization = x, quiet = TRUE)
      TRUE
    }, error = function(e) FALSE)
    if (ok) return(x)
  }
  stop("Could not find a supported gmrf_parameterization value for dsem")
}

# Apply stored centering and scaling constants to new data.
scale_with <- function(x, center, scale) {
  sweep(sweep(x, 2, center, "-"), 2, scale, "/")
}

# Return the last finite value in a numeric vector.
last_finite <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) == 0) NA_real_ else tail(x, 1)
}

# Fit a regional DSEM, compute leave-one-out diagnostics, and generate an
# illustrative forward TAC forecast from observed or supplied ABC scenarios.
fit_dsem_region <- function(
  data,
  dep_species,
  indep_species,
  n_future = 5,
  seed = 1,
  future_abc_raw = NULL,
  future_years = NULL,
  compute_loo = TRUE
) {
  dsem_df <- data %>%
    filter(Species %in% unique(c(dep_species, indep_species))) %>%
    group_by(Year, Species) %>%
    summarise(
      TAC = sum(TAC, na.rm = TRUE),
      ABC = sum(ABC, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from = Species,
      values_from = c(TAC, ABC),
      names_sep = "_"
    ) %>%
    arrange(Year)

  names(dsem_df) <- safe_names(names(dsem_df))

  dep_safe <- safe_names(dep_species)
  indep_safe <- safe_names(indep_species)
  y_cols <- paste0("TAC_", dep_safe)
  x_cols <- paste0("ABC_", indep_safe)

  ts_raw <- dsem_df[, c(y_cols, x_cols)]
  ts_scaled <- scale(ts_raw)
  ts_center <- attr(ts_scaled, "scaled:center")
  ts_scale <- attr(ts_scaled, "scaled:scale")
  ts_center[is.na(ts_center)] <- 0
  ts_scale[is.na(ts_scale) | ts_scale == 0] <- 1

  tsdata <- ts(ts_scaled, start = min(dsem_df$Year))

  ar_lines <- paste0(y_cols, " -> ", y_cols, ", 1, ar_", dep_safe)
  abc_lines <- unlist(lapply(y_cols, function(y) {
    paste0(x_cols, " -> ", y, ", 0, b_", gsub("ABC_", "", x_cols), "_to_", gsub("TAC_", "", y))
  }))
  sem <- paste(c(ar_lines, abc_lines), collapse = "\n")

  gmrf_param <- choose_gmrf_param()

  # Try one specific DSEM configuration so callers can loop over fallbacks.
  fit_dsem_once <- function(gmrf, family_opt = NULL) {
    args <- list(
      sem = sem,
      tsdata = tsdata,
      estimate_delta0 = TRUE,
      control = dsem_control(
        quiet = TRUE,
        gmrf_parameterization = gmrf,
        newton_loops = 0,
        extra_convergence_checks = FALSE,
        getsd = TRUE
      )
    )
    if (!is.null(family_opt)) args$family <- family_opt
    do.call(dsem::dsem, args)
  }

  gmrf_candidates <- unique(c(
    gmrf_param,
    "full", "project", "gmrf_project", "mvn_project", "separable", "projection"
  ))

  fit <- NULL
  fit_notes <- character()

  for (g in gmrf_candidates) {
    for (fam in list(NULL, "gaussian")) {
      fam_tag <- if (is.null(fam)) "default" else fam
      res <- tryCatch(
        fit_dsem_once(g, fam),
        error = function(e) e
      )
      if (!inherits(res, "error")) {
        fit <- res
        fit_notes <- c(fit_notes, paste0("fit succeeded with gmrf=", g, ", family=", fam_tag))
        break
      } else {
        fit_notes <- c(
          fit_notes,
          paste0("fit failed with gmrf=", g, ", family=", fam_tag, ": ", conditionMessage(res))
        )
      }
    }
    if (!is.null(fit)) break
  }

  if (is.null(fit)) {
    stop(
      paste(
        "DSEM fit failed for all gmrf/family combinations.",
        paste(fit_notes, collapse = " | ")
      )
    )
  }

  dsem_summary <- summary(fit)
  if (isTRUE(compute_loo)) {
    loo_tbl <- dsem::loo_residuals(fit, what = "loo", track_progress = FALSE) %>%
      as_tibble() %>%
      rename(
        Year = 1,
        Series = 2,
        Observed_scaled = obs,
        Predicted_scaled = est,
        SE_scaled = se
      ) %>%
      mutate(
        Year = as.numeric(Year),
        Series = as.character(Series)
      )
  } else {
    loo_tbl <- tibble(
      Year = numeric(),
      Series = character(),
      Observed_scaled = numeric(),
      Predicted_scaled = numeric(),
      SE_scaled = numeric()
    )
  }

  if (is.null(future_abc_raw)) {
    set.seed(seed)

    last_abc <- vapply(dsem_df[x_cols], last_finite, numeric(1))
    growth <- setNames(rep(-0.005, length(x_cols)), x_cols)
    dep_abc_cols <- intersect(paste0("ABC_", dep_safe), x_cols)
    if (length(dep_abc_cols) >= 1) growth[dep_abc_cols[1]] <- 0.02
    if (length(dep_abc_cols) >= 2) growth[dep_abc_cols[2]] <- 0.01

    noise_sd <- 0.03

    new_abc_raw <- sapply(seq_len(n_future), function(h) {
      eps <- rnorm(length(x_cols), mean = 0, sd = noise_sd)
      pmax(last_abc * (1 + growth) ^ h * (1 + eps), 1)
    })
    new_abc_raw <- t(new_abc_raw)
    colnames(new_abc_raw) <- x_cols
    forecast_years <- max(dsem_df$Year) + seq_len(n_future)
  } else {
    new_abc_raw <- future_abc_raw %>%
      as.data.frame() %>%
      select(all_of(x_cols))
    n_future <- nrow(new_abc_raw)
    forecast_years <- if (is.null(future_years)) {
      max(dsem_df$Year) + seq_len(n_future)
    } else {
      future_years
    }
  }

  new_abc_scaled <- scale_with(new_abc_raw, ts_center[x_cols], ts_scale[x_cols])

  newdata <- as.data.frame(tsdata)
  newdata_future <- as.data.frame(new_abc_scaled)
  newdata_future[, y_cols] <- NA_real_
  newdata <- bind_rows(newdata, newdata_future)
  newdata <- newdata[, colnames(as.data.frame(tsdata))]

  newdata_ts <- ts(as.matrix(newdata), start = start(tsdata), frequency = frequency(tsdata))

  pred_scaled <- predict(fit, newdata = newdata_ts, type = "response")
  colnames(pred_scaled) <- colnames(newdata_ts)
  pred <- sweep(pred_scaled, 2, ts_scale[colnames(pred_scaled)], "*")
  pred <- sweep(pred, 2, ts_center[colnames(pred)], "+")

  pred_tac_future <- tail(pred[, y_cols], n_future)
  forecast_tbl <- as_tibble(new_abc_raw) %>%
    mutate(Year = forecast_years) %>%
    bind_cols(as_tibble(pred_tac_future))

  list(
    fit = fit,
    summary = dsem_summary,
    dsem_df = dsem_df,
    dep_species = dep_species,
    y_cols = y_cols,
    x_cols = x_cols,
    pred = pred,
    ts_center = ts_center,
    ts_scale = ts_scale,
    loo_tbl = loo_tbl,
    n_future = n_future,
    forecast_years = forecast_years,
    forecast_tbl = forecast_tbl,
    fit_notes = fit_notes
  )
}

# Refit the DSEM sequentially through time and score one-step-ahead forecasts.
run_dsem_retrospective <- function(data, dep_species, indep_species, start_year = 2015, seed = 1) {
  full_df <- data %>%
    filter(Species %in% unique(c(dep_species, indep_species))) %>%
    group_by(Year, Species) %>%
    summarise(
      TAC = sum(TAC, na.rm = TRUE),
      ABC = sum(ABC, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from = Species,
      values_from = c(TAC, ABC),
      names_sep = "_"
    ) %>%
    arrange(Year)

  names(full_df) <- safe_names(names(full_df))

  dep_safe <- safe_names(dep_species)
  indep_safe <- safe_names(indep_species)
  y_cols <- paste0("TAC_", dep_safe)
  x_cols <- paste0("ABC_", indep_safe)

  target_years <- full_df %>%
    filter(Year >= start_year) %>%
    pull(Year)

  purrr::map_dfr(seq_along(target_years), function(i) {
    target_year <- target_years[[i]]
    target_row <- full_df %>%
      filter(Year == target_year)

    if (nrow(target_row) != 1 || any(!is.finite(unlist(target_row[, x_cols, drop = FALSE])))) {
      return(tibble(
        Year = target_year,
        Training_end = target_year - 1,
        Species = dep_species,
        Observed = as.numeric(target_row[, y_cols, drop = FALSE]),
        Predicted = NA_real_,
        Error = NA_real_,
        Abs_Error = NA_real_,
        APE = NA_real_,
        Status = "missing_future_abc"
      ))
    }

    fit_i <- tryCatch(
      fit_dsem_region(
        data = data %>% filter(Year < target_year),
        dep_species = dep_species,
        indep_species = indep_species,
        n_future = 1,
        seed = seed + i,
        future_abc_raw = target_row[, x_cols, drop = FALSE],
        future_years = target_year,
        compute_loo = FALSE
      ),
      error = function(e) e
    )

    if (inherits(fit_i, "error")) {
      return(tibble(
        Year = target_year,
        Training_end = target_year - 1,
        Species = dep_species,
        Observed = as.numeric(target_row[, y_cols, drop = FALSE]),
        Predicted = NA_real_,
        Error = NA_real_,
        Abs_Error = NA_real_,
        APE = NA_real_,
        Status = paste0("fit_failed: ", conditionMessage(fit_i))
      ))
    }

    predicted <- fit_i$forecast_tbl %>%
      select(all_of(y_cols))

    tibble(
      Year = target_year,
      Training_end = target_year - 1,
      Species = dep_species,
      Observed = as.numeric(target_row[, y_cols, drop = FALSE]),
      Predicted = as.numeric(predicted[1, , drop = FALSE]),
      Status = "ok"
    ) %>%
      mutate(
        Error = Observed - Predicted,
        Abs_Error = abs(Error),
        APE = if_else(Observed != 0, 100 * Abs_Error / abs(Observed), NA_real_)
      )
  })
}

# Re-label raw DSEM coefficient output into management-facing term groups.
summarize_dsem_results <- function(dsem_summary) {
  dsem_summary %>%
    mutate(
      Term = if_else(lag > 0, paste0(path, " (lag ", lag, ")"), path),
      Group = case_when(
        lag > 0 ~ "Autoregressive (lagged TAC)",
        str_detect(first, "^ABC_") ~ "ABC -> TAC",
        TRUE ~ "Other"
      )
    ) %>%
    arrange(Group, Term) %>%
    select(Group, Term, Estimate, Std_Error, z_value, p_value)
}

# Render the future ABC inputs and forecast TAC outputs as a gt table.
render_dsem_forecast_table <- function(region_fit) {
  region_fit$forecast_tbl %>%
    select(Year, starts_with("ABC_"), starts_with("TAC_")) %>%
    gt::gt() %>%
    gt::fmt_number(columns = starts_with("ABC_"), decimals = 0) %>%
    gt::fmt_number(columns = starts_with("TAC_"), decimals = 0)
}

# Render the grouped DSEM coefficient summary as a gt table.
render_dsem_summary_table <- function(region_fit) {
  summarize_dsem_results(region_fit$summary) %>%
    gt::gt(rowname_col = "Term", groupname_col = "Group") %>%
    gt::cols_label(
      Estimate = "Estimate",
      Std_Error = "Std. Error",
      z_value = "z",
      p_value = "p"
    ) %>%
    gt::fmt_number(columns = c(Estimate, Std_Error), decimals = 3) %>%
    gt::fmt_number(columns = z_value, decimals = 2) %>%
    gt::fmt_number(columns = p_value, decimals = 3)
}

# Back-transform leave-one-out residuals into TAC units for one fitted region.
extract_dsem_loo_residuals <- function(region_fit, region_name) {
  region_fit$loo_tbl %>%
    filter(Series %in% region_fit$y_cols) %>%
    mutate(
      Region = region_name,
      Species = region_fit$dep_species[match(Series, region_fit$y_cols)],
      Observed = Observed_scaled * region_fit$ts_scale[Series] + region_fit$ts_center[Series],
      Predicted = Predicted_scaled * region_fit$ts_scale[Series] + region_fit$ts_center[Series],
      SE = SE_scaled * region_fit$ts_scale[Series],
      Residual = Observed - Predicted,
      Std_Residual = if_else(SE_scaled > 0, (Observed_scaled - Predicted_scaled) / SE_scaled, NA_real_)
    ) %>%
    select(Region, Year, Species, Observed, Predicted, SE, Residual, Std_Residual)
}

# Bind leave-one-out residual diagnostics across all fitted regions.
collect_dsem_loo_residuals <- function(region_fits) {
  purrr::imap_dfr(region_fits, extract_dsem_loo_residuals)
}

# Summarize leave-one-out residual performance metrics by region and species.
render_dsem_residual_comparison_table <- function(region_fits) {
  collect_dsem_loo_residuals(region_fits) %>%
    group_by(Region, Species) %>%
    summarise(
      Bias = mean(Residual, na.rm = TRUE),
      MAE = mean(abs(Residual), na.rm = TRUE),
      RMSE = sqrt(mean(Residual^2, na.rm = TRUE)),
      Mean_abs_z = mean(abs(Std_Residual), na.rm = TRUE),
      Max_abs_z = max(abs(Std_Residual), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(Region, Species) %>%
    gt::gt(rowname_col = "Species", groupname_col = "Region") %>%
    gt::cols_label(
      Bias = "Bias",
      MAE = "MAE",
      RMSE = "RMSE",
      Mean_abs_z = "Mean |z|",
      Max_abs_z = "Max |z|"
    ) %>%
    gt::fmt_number(columns = c(Bias, MAE, RMSE), decimals = 0) %>%
    gt::fmt_number(columns = c(Mean_abs_z, Max_abs_z), decimals = 2)
}

# Plot standardized leave-one-out residuals through time by species and region.
plot_dsem_loo_residuals <- function(region_fits) {
  plot_df <- collect_dsem_loo_residuals(region_fits)

  ggplot(plot_df, aes(x = Year, y = Std_Residual)) +
    geom_hline(yintercept = 0, color = "grey60", linewidth = 0.4) +
    geom_hline(yintercept = c(-2, 2), color = "firebrick", linetype = "dashed", linewidth = 0.4) +
    geom_line(color = "steelblue", linewidth = 0.8) +
    geom_point(color = "steelblue", size = 1.5, alpha = 0.85) +
    facet_grid(Species ~ Region) +
    labs(x = NULL, y = "Standardized leave-one-out residual")
}

# Combine retrospective forecast results from each region into one data frame.
collect_dsem_retrospective <- function(retro_fits) {
  purrr::imap_dfr(retro_fits, ~ mutate(.x, Region = .y), .id = NULL)
}

# Summarize one-step-ahead retrospective forecast accuracy by region and species.
render_dsem_retrospective_summary_table <- function(retro_fits) {
  collect_dsem_retrospective(retro_fits) %>%
    filter(Status == "ok") %>%
    group_by(Region, Species) %>%
    summarise(
      n_years = n(),
      Bias = mean(Error, na.rm = TRUE),
      MAE = mean(Abs_Error, na.rm = TRUE),
      RMSE = sqrt(mean(Error^2, na.rm = TRUE)),
      MAPE = mean(APE, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(Region, Species) %>%
    gt::gt(rowname_col = "Species", groupname_col = "Region") %>%
    gt::cols_label(
      n_years = "Years",
      Bias = "Bias",
      MAE = "MAE",
      RMSE = "RMSE",
      MAPE = "MAPE (%)"
    ) %>%
    gt::fmt_number(columns = n_years, decimals = 0) %>%
    gt::fmt_number(columns = c(Bias, MAE, RMSE), decimals = 0) %>%
    gt::fmt_number(columns = MAPE, decimals = 2)
}

# Plot observed versus retrospective one-step-ahead TAC predictions.
plot_dsem_retrospective <- function(retro_fits) {
  plot_df <- collect_dsem_retrospective(retro_fits) %>%
    filter(Status == "ok") %>%
    pivot_longer(c(Observed, Predicted), names_to = "Series", values_to = "TAC")

  ggplot(plot_df, aes(x = Year, y = TAC, color = Series)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 1.5, alpha = 0.85) +
    facet_grid(Species ~ Region, scales = "free_y") +
    scale_color_manual(values = c(Observed = "grey40", Predicted = "steelblue")) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "TAC", color = NULL)
}

# Plot in-sample fitted TAC values plus the forward forecast extension.
plot_dsem_predictions <- function(region_fit) {
  pred_years <- c(region_fit$dsem_df$Year, region_fit$forecast_years)
  pred_df <- as_tibble(region_fit$pred[, region_fit$y_cols]) %>%
    mutate(Year = pred_years) %>%
    pivot_longer(-Year, names_to = "Species", values_to = "Predicted")

  obs_df <- region_fit$dsem_df %>%
    select(Year, all_of(region_fit$y_cols)) %>%
    pivot_longer(-Year, names_to = "Species", values_to = "Observed")

  plot_df <- left_join(pred_df, obs_df, by = c("Year", "Species"))

  ggplot(plot_df, aes(x = Year)) +
    geom_line(aes(y = Predicted, color = "Predicted"), linewidth = 1) +
    geom_point(aes(y = Observed, color = "Observed"), size = 1.4, alpha = 0.8) +
    facet_grid(Species ~ ., scales = "free_y") +
    scale_color_manual(values = c(Predicted = "steelblue", Observed = "grey40")) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "TAC", color = NULL)
}

bsai_dsem_dep <- c("Pollock", "Pacific cod")
bsai_dsem_indep <- mainspp
bsai_dsem <- fit_dsem_region(
  data = dfOY %>% filter(lag == 1, Species %in% mainspp),
  dep_species = bsai_dsem_dep,
  indep_species = bsai_dsem_indep
)

goa_dsem_dep <- c("Pollock", "Pacific cod")
goa_dsem_indep <- goaraw %>%
  filter(lag == 1, OY == 1, Area == "Total") %>%
  group_by(Species) %>%
  summarise(total_TAC = sum(TAC, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(total_TAC), total_TAC > 0) %>%
  arrange(desc(total_TAC)) %>%
  slice_head(n = 7) %>%
  pull(Species)

goa_dsem <- fit_dsem_region(
  data = goaOY %>% filter(lag == 1, Species %in% goa_dsem_indep),
  dep_species = goa_dsem_dep,
  indep_species = goa_dsem_indep
)

bsai_dsem_retro <- run_dsem_retrospective(
  data = dfOY %>% filter(lag == 1, Species %in% mainspp),
  dep_species = bsai_dsem_dep,
  indep_species = bsai_dsem_indep,
  start_year = 2015
)

goa_dsem_retro <- run_dsem_retrospective(
  data = goaOY %>% filter(lag == 1, Species %in% goa_dsem_indep),
  dep_species = goa_dsem_dep,
  indep_species = goa_dsem_indep,
  start_year = 2015
)
```

### BSAI

```{r}
#| label: fig-dsem-dag-contemporaneous-bsai
#| fig-cap: "BSAI contemporaneous DSEM DAG with ABC drivers for Pollock and Pacific cod TACs."
plot_dsem_dag(
  dep_species = bsai_dsem_dep,
  indep_species = setdiff(bsai_dsem_indep, bsai_dsem_dep),
  dynamic = FALSE
)
```

```{r}
#| label: fig-dsem-dag-dynamic-bsai
#| fig-cap: "BSAI dynamic DSEM DAG with AR(1) TAC structure and contemporaneous ABC effects."
plot_dsem_dag(
  dep_species = bsai_dsem_dep,
  indep_species = setdiff(bsai_dsem_indep, bsai_dsem_dep),
  dynamic = TRUE
)
```

### GOA

```{r}
#| label: fig-dsem-dag-contemporaneous-goa
#| fig-cap: "GOA contemporaneous DSEM DAG with ABC drivers for Pollock and Pacific cod TACs."
plot_dsem_dag(
  dep_species = goa_dsem_dep,
  indep_species = setdiff(goa_dsem_indep, goa_dsem_dep),
  dynamic = FALSE
)
```

```{r}
#| label: fig-dsem-dag-dynamic-goa
#| fig-cap: "GOA dynamic DSEM DAG with AR(1) TAC structure and contemporaneous ABC effects."
plot_dsem_dag(
  dep_species = goa_dsem_dep,
  indep_species = setdiff(goa_dsem_indep, goa_dsem_dep),
  dynamic = TRUE
)
```

## DSEM estimates
In both regions, the DSEM estimates TAC dynamics for Pollock and Pacific cod with two components: (1) AR(1) persistence in each TAC time series and (2) contemporaneous influence of ABC inputs, including both focal-stock ABC and cross-stock ABCs. Models are fit on scaled inputs for stability, then predictions are transformed back to the original TAC units for interpretation. The forecast tables below remain illustrative and use the same simple forward ABC scenario in each region.

We also compare leave-one-out residuals from `dsem::loo_residuals()` for the fitted TAC series. These diagnostics show whether the dynamic model leaves systematic bias or large standardized errors for Pollock and Pacific cod in either region when each year is predicted from the rest of the time series.

### BSAI

For BSAI, the DSEM uses the seven main stocks (`r paste(bsai_dsem_indep, collapse = ", ")`) as contemporaneous ABC inputs.

```{r}
#| label: tbl-dsem-forecast-bsai
#| tbl-cap: "BSAI example DSEM forecast: assumed future ABCs and predicted TACs (illustrative scenario)."
render_dsem_forecast_table(bsai_dsem)
```

```{r}
#| label: tbl-dsem-summary-bsai
#| tbl-cap: "Summary of BSAI DSEM results."
render_dsem_summary_table(bsai_dsem)
```

```{r}
#| label: fig-dsem-pred-bsai
#| fig-cap: "Observed vs predicted BSAI TACs from the DSEM (back-transformed to original units)."
#| fig-height: 5
#| fig-width: 7
plot_dsem_predictions(bsai_dsem)
```

### GOA

For GOA, the DSEM uses the top seven GOA species by total TAC from the `Area == "Total"` series (`r paste(goa_dsem_indep, collapse = ", ")`) as contemporaneous ABC inputs.

```{r}
#| label: tbl-dsem-forecast-goa
#| tbl-cap: "GOA example DSEM forecast: assumed future ABCs and predicted TACs (illustrative scenario)."
render_dsem_forecast_table(goa_dsem)
```

```{r}
#| label: tbl-dsem-summary-goa
#| tbl-cap: "Summary of GOA DSEM results."
render_dsem_summary_table(goa_dsem)
```

```{r}
#| label: fig-dsem-pred-goa
#| fig-cap: "Observed vs predicted GOA TACs from the DSEM (back-transformed to original units)."
#| fig-height: 5
#| fig-width: 7
plot_dsem_predictions(goa_dsem)
```

### Leave-one-out residual comparisons

```{r}
#| label: tbl-dsem-loo-residuals
#| tbl-cap: "Leave-one-out TAC residual summaries for the BSAI and GOA DSEM fits."
render_dsem_residual_comparison_table(list(BSAI = bsai_dsem, GOA = goa_dsem))
```

```{r}
#| label: fig-dsem-loo-residuals
#| fig-cap: "Standardized leave-one-out TAC residuals for the BSAI and GOA DSEM fits."
#| fig-height: 5.5
#| fig-width: 8
plot_dsem_loo_residuals(list(BSAI = bsai_dsem, GOA = goa_dsem))
```

### Retrospective skill test from 2015 onward

For each year from 2015 onward, we refit the DSEM using only data available through the previous year and then forecast that year’s Pollock and Pacific cod TACs using the realized ABC vector for that year. This is a rolling-origin, leave-future-out test of one-step-ahead skill rather than an in-sample fit diagnostic.

```{r}
#| label: tbl-dsem-retrospective
#| tbl-cap: "Rolling one-step-ahead DSEM skill from 2015 onward, using only prior history to predict each year."
render_dsem_retrospective_summary_table(list(BSAI = bsai_dsem_retro, GOA = goa_dsem_retro))
```

```{r}
#| label: fig-dsem-retrospective
#| fig-cap: "Rolling one-step-ahead DSEM forecasts versus observed TAC from 2015 onward."
#| fig-height: 5.5
#| fig-width: 8
plot_dsem_retrospective(list(BSAI = bsai_dsem_retro, GOA = goa_dsem_retro))
```


# Discussion
## Projection updates under new data and assessment

Multi-year catch advice inevitably embeds a forecasting problem: projections are conditioned on assumptions about future recruitment, selectivity, mortality, and implementation error, all frozen at the time advice is issued [@MAFMC2012MultiYearABC; @ICESAdviceFramework]. The literature above converges on a shared observation: when a new assessment assimilates additional data, the posterior state of the stock can shift abruptly relative to those earlier projections [@SanchezMarono2021Lag]. The impact of updating data and assessment 
creates a  discontinuity and reflects the consequence of evolving data streams and analyses.

Interim management procedure (MP) work reframes the problem by explicitly separating assessment cadence from advice cadence [@Huynh2020InterimMP; @Klibansky2022InterimAdvice]. Rather than treating interim updates as ad-hoc corrections, these approaches formalize how limited new information (recent catches, survey indices, environmental indicators) should update catch advice between full assessments. The goal is not to eliminate snap—information shocks are unavoidable—but to bound it, so that revisions are predictable, explainable, and risk-consistent.

This framing is directly relevant to BSAI groundfish, where large stocks (pollock, cod, flatfish) dominate total ABC and where management stability is highly valued. Multi-year ABCs smooth year-to-year variability, but smoothing pushes uncertainty forward in time. When a subsequent assessment revises recruitment trajectories or selectivity patterns, the resulting adjustment can appear disproportionate relative to the modest interim changes that preceded it. The Canadian and U.S. guidance documents emphasize that this tension is intrinsic: stability and responsiveness trade off, and neither can be maximized simultaneously [@Krohn2019DFOInterim; @MAFMC2012MultiYearABC; @SEDAR2022Projections].

The ICES and MSE-oriented literature adds an important synthesis [@ICESAdviceFramework]. Geromont and Butterworth [@Geromont2015ComplexVsSimple] and Walter et al. [@Walter2023WhenMSE] show that systems relying on simple, pre-tested update rules often experience smaller and more interpretable snaps than systems that defer all learning to infrequent, high-dimensional assessments. In that sense, forecast-to-assimilation snap is best understood as a design property of the advice system, not merely an outcome of assessment error. For BSAI groundfish, this perspective motivates explicit evaluation of how multi-year projections, rollover rules, and interim updates distribute learning over time—shaping not only biological risk, but also the credibility and usability of scientific advice.


# Appendix
## Function catalog

This appendix documents the custom functions defined in this report. The two `safe_names()` definitions are intentionally duplicated so the regression and DSEM chunks can run independently during rendering.

```{r}
#| label: tbl-function-catalog
#| tbl-cap: "Catalog of custom functions defined in the report."
function_catalog <- tibble::tribble(
  ~Workflow, ~Function, ~Scope, ~Purpose,
  "Exploratory plots", "lower_points_smooth", "top-level", "Helper for `GGally::ggpairs()` lower panels with points plus a loess smooth.",
  "Regression", "safe_names", "top-level", "Sanitizes species and column names into ASCII-safe identifiers for regression design matrices.",
  "Regression", "fit_tac_abc_regressions", "top-level", "Builds grouped TAC/ABC data, fits stock-specific log-linear regressions, and returns coefficient and prediction objects.",
  "Regression", "render_regression_table", "top-level", "Formats regression coefficients and fit statistics with `gt`.",
  "Regression", "plot_regression_predictions", "top-level", "Plots observed versus predicted TAC trajectories from the regression fits.",
  "GOA OY cap", "rescale_oy_cap", "top-level", "Implements weighted cap rescaling and returns the rescaled catch vector plus the solved multiplier.",
  "GOA OY cap", "total_gap", "nested in `rescale_oy_cap()`", "Objective helper used in the bisection search to match the aggregate OY cap.",
  "DSEM", "safe_names", "top-level", "Sanitizes species and state names used to build DSEM series names and path labels.",
  "DSEM", "centered_positions", "top-level", "Creates vertically centered node positions for static and dynamic DAG plots.",
  "DSEM", "plot_dsem_dag", "top-level", "Builds static or dynamic DAG visualizations of the DSEM structure with `ggdag`.",
  "DSEM", "choose_gmrf_param", "top-level", "Finds a `gmrf_parameterization` value accepted by the installed `dsem` version.",
  "DSEM", "scale_with", "top-level", "Applies stored centering and scaling constants to new predictor matrices.",
  "DSEM", "last_finite", "top-level", "Returns the last finite value in a series, used to seed the forward ABC scenario.",
  "DSEM", "fit_dsem_region", "top-level", "Prepares regional multivariate time series, fits the DSEM, computes forecasts, and caches leave-one-out residual output.",
  "DSEM", "fit_dsem_once", "nested in `fit_dsem_region()`", "Small wrapper that calls `dsem::dsem()` for one GMRF/family combination.",
  "DSEM skill testing", "run_dsem_retrospective", "top-level", "Runs a rolling one-step-ahead retrospective from 2015 onward by fitting each target year only on prior history.",
  "DSEM", "summarize_dsem_results", "top-level", "Re-labels and groups `summary(fit)` output into interpretable DSEM coefficient categories.",
  "DSEM", "render_dsem_forecast_table", "top-level", "Formats the illustrative future ABC/TAC scenario table.",
  "DSEM", "render_dsem_summary_table", "top-level", "Formats the DSEM coefficient summary table with `gt`.",
  "DSEM residuals", "extract_dsem_loo_residuals", "top-level", "Back-transforms TAC leave-one-out predictions and residuals for one regional fit.",
  "DSEM residuals", "collect_dsem_loo_residuals", "top-level", "Combines leave-one-out residual data across multiple regional DSEM fits.",
  "DSEM residuals", "render_dsem_residual_comparison_table", "top-level", "Summarizes bias, MAE, RMSE, and standardized residual magnitudes across regions and species.",
  "DSEM residuals", "plot_dsem_loo_residuals", "top-level", "Plots standardized leave-one-out TAC residuals through time.",
  "DSEM skill testing", "collect_dsem_retrospective", "top-level", "Combines rolling one-step-ahead results across regional retrospective runs.",
  "DSEM skill testing", "render_dsem_retrospective_summary_table", "top-level", "Summarizes rolling retrospective skill using bias, MAE, RMSE, and MAPE.",
  "DSEM skill testing", "plot_dsem_retrospective", "top-level", "Plots observed TAC against rolling one-step-ahead DSEM forecasts from 2015 onward.",
  "DSEM", "plot_dsem_predictions", "top-level", "Plots observed and predicted TAC values from fitted and forecast DSEM output."
)

function_catalog %>%
  gt::gt(rowname_col = "Function", groupname_col = "Workflow") %>%
  gt::cols_label(
    Scope = "Scope",
    Purpose = "Purpose"
  ) %>%
  gt::cols_width(
    Scope ~ gt::px(180),
    Purpose ~ gt::px(540)
  )
```

## DSEM package specifics

The DSEM portion of the report depends on package-specific behavior in `dsem`, especially around model specification, version compatibility, prediction, and residual diagnostics. The table below records the main package interfaces and how they are used here.

```{r}
#| label: tbl-dsem-package-notes
#| tbl-cap: "DSEM package interfaces and implementation details used in this report."
dsem_package_notes <- tibble::tribble(
  ~Topic, ~Package_interface, ~How_used_here, ~Why_it_matters,
  "Model fitting", "`dsem::dsem()`", "Fits each regional multivariate TAC-ABC time series after converting the data frame to a scaled `ts` object.", "This is the core model fit that produces latent states, coefficient summaries, and predictions.",
  "Version compatibility", "`dsem::dsem_control(gmrf_parameterization = ...)`", "The helper `choose_gmrf_param()` tests several candidate values because accepted names differ across installed package versions.", "Without this compatibility layer, the same report can fail locally or in CI when `dsem` changes its control vocabulary.",
  "SEM specification", "line-based SEM string", "Autoregressive TAC paths and contemporaneous ABC-to-TAC paths are concatenated into the `sem` string passed to `dsem()`.", "The report constructs the model explicitly rather than relying on a canned template, so the path structure is transparent in the source.",
  "State initialization", "`estimate_delta0 = TRUE`", "Enabled for both BSAI and GOA fits inside `fit_dsem_region()`.", "This lets the model estimate initial state offsets rather than forcing a fixed starting value.",
  "Standardization", "base `scale()` plus stored centers/scales", "All TAC and ABC series are fit on a common scaled basis, then predictions and leave-one-out outputs are transformed back to TAC units for display.", "Scaling improves optimizer stability and makes coefficients comparable across series with very different magnitudes.",
  "Prediction", "`predict(fit, newdata = ..., type = \"response\")`", "Used after appending future ABC scenarios with missing TAC values to obtain fitted-plus-forecast TAC paths.", "This provides a single prediction pass covering observed years and the illustrative forecast horizon.",
  "Retrospective skill", "rolling one-step-ahead refits", "For each year from 2015 onward, the report refits the DSEM on data through `year - 1` and predicts that year using the realized ABC vector.", "This is the leave-future-out skill test in the DSEM section and gives a stricter assessment than full-sample fitted values.",
  "Fallback fit strategy", "candidate GMRF and family loop", "The code tries multiple GMRF parameterizations and both default and Gaussian family settings before failing.", "This makes the document more robust to package and platform differences without changing the scientific specification.",
  "Coefficient summaries", "`summary(fit)`", "The raw summary output is regrouped into autoregressive and ABC-to-TAC terms before rendering.", "The package summary is machine-oriented; the report labels it in management-facing terms.",
  "LOO residual diagnostics", "`dsem::loo_residuals(fit, what = \"loo\")`", "The report uses the returned table of observed values, leave-one-out estimates, and standard errors, then back-transforms TAC series and computes standardized residuals.", "This provides a package-native cross-check on fit quality that is more stringent than in-sample residual inspection.",
  "Residual scale handling", "`obs`, `est`, and `se` columns from `loo_residuals()`", "Residual summaries are computed as TAC-unit errors after back-transformation, while standardized residuals remain on the scaled latent-state basis.", "Keeping both scales separates management-scale error magnitude from model-based z-score diagnostics."
)

dsem_package_notes %>%
  gt::gt(rowname_col = "Topic") %>%
  gt::cols_label(
    Package_interface = "Package interface",
    How_used_here = "How used here",
    Why_it_matters = "Why it matters"
  ) %>%
  gt::cols_width(
    Package_interface ~ gt::px(180),
    How_used_here ~ gt::px(300),
    Why_it_matters ~ gt::px(300)
  )
```

## DSEM object contents used by the report

For the regional DSEM fits, the report retains a small structured list rather than passing the raw `dsem` fit object through every downstream chunk. The most important retained components are:

- `fit`: the original fitted `dsem` object, needed for any additional package-native inspection.
- `summary`: the coefficient summary returned by `summary(fit)`.
- `dsem_df`, `y_cols`, and `x_cols`: the regional data frame and the TAC/ABC series names used to build the model.
- `pred`, `forecast_years`, and `forecast_tbl`: fitted values plus the illustrative forward ABC scenario and resulting TAC forecasts.
- `ts_center` and `ts_scale`: the centering and scaling constants required to back-transform predictions and leave-one-out residuals.
- `loo_tbl`: the leave-one-out residual table returned by `dsem::loo_residuals(..., what = "loo")`.
- `fit_notes`: a log of which GMRF/family combinations failed or succeeded during fitting.


# References
